---
title: "Tidyverse & ggplot2 - ICD Bootcamp"
author: "Syed Murtuza Baker"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    theme: flatly
    toc: yes
    number_sections: true
    toc_float:
      collapsed: true
      smooth_scroll: true
editor_options: 
  chunk_output_type: inline
---



# Tidyverse
tidyverse.org defines Tidyverse as

> The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures.

```{r}
library(tidyverse)
library(dplyr)
library(scater)
```

We will use single-cell RNA sequencing on 6826 stem cells from Chronic myelomonocytic leukaemia (CMML) patients and healthy controls using the droplet-based, ultra-high-throughput 10x platform. We found substantial inter and intra-patient heterogeneity, with CMML stem cells displaying distinctive transcriptional programs. Compared with normal controls, CMML stem cells exhibited transcriptomes characterized by increased expression of myeloid-lineage and cell cycle genes, and lower expression of genes selectively expressed by normal haematopoietic stem cells. 

```{r}
sce <- readRDS('sce.rds')
sce
```

## Tibble
- Tibbles are data-frames
- `tibble()` does much less: it never changes the type of the inputs (e.g. it never converts strings to factors!)
- it never changes the names of variables, and it never creates row names.
- tibble can have column names that are not valid R variable names, aka *non-syntactic* names.
```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb
```

## Pipe `%>%`
Pipe `%>%` passes the output from one stage to the other.
```{r}
tbl_df(colData(sce))
```
```{r}
names(colData(sce))
```


```{r}
tbl_df(colData(sce)) %>%
  group_by(Sample) %>%
  summarise(
    total.features = mean(total_features),
    total.counts = mean(total_counts)
  )
```

## `dplyr` - Functions as verbs.

__The most useful__

- `select()`: select columns
- `mutate()`: create new variables, change existing
- `filter()`: subset your data by some criterion
- `summarize()`: summarize your data in some way
- `group_by()`: group your data by a variable
- `slice()`: grab specific rows
- `select()`: select an observation

__Some others__

- `count()`: count your data
- `arrange()`: arrange your data by a column or variable
- `distinct()`: gather all distinct values of a variable
- `n_distinct()`: count how many distinct values you have (only works with summarize)
- `n()`: count how many observation you have for a subgroup
- `sample_n()`: Grab an N sample of your data
- `ungroup()`: ungroup grouped data by a variable
- `top_n(`): get the top N number of entries from a data frame

__To make it easier we copy the metadata for our `SingleCellExperiment` object `sce` to `d` __

```{r}
d <- tbl_df(colData(sce))
```

### `Select` : To select collumns
```{r}
select(d, Sample, Cluster, cellType)
```

```{r}
d %>% 
  select(Sample, Cluster, cellType)
```

### `Filter` : To select rows
```{r}
d %>% 
  filter(cellType == "HSC")
```

```{r}
d %>% 
  select(barcode, Sample, total_features, cellType, Cluster) %>%
  filter(Sample == "BC572")

```

```{r}
d %>% 
  filter(cellType == "Erythrocytes", pct_counts_Mt > 1.5) %>% 
  select(barcode, Sample, pct_counts_Mt, cellType, Cluster)
```

### `Mutate`: 
To create new variables in the data table:
```{r}
d_exp <- d
d_exp <- cbind(d_exp, t(logcounts(sce)[c('KLF4','RUNX1','EGR1'),]))
```

```{r}
d_exp
```

```{r}
d_exp %>% 
  mutate(Klf4Diff = abs(KLF4 - RUNX1)) %>%
  select(barcode, Sample, cellType, Klf4Diff)
```



### `Arrange`: 
To order the data by a particular variable:

```{r}
d_exp %>% 
  mutate(Klf4Diff = abs(KLF4 - RUNX1)) %>% 
  arrange(desc(Klf4Diff)) %>% 
  slice(1)
```
### `Slice`: 
To slice your data by rows:

```{r}
# The top 5 goleadas?
d_exp %>% 
  mutate(Klf4Diff = abs(KLF4 - RUNX1)) %>% 
  arrange(desc(Klf4Diff)) %>% 
  slice(1:5)  # slice_max here would also do the trick
```

```{r}
# The top 5 goleadas?
d_exp %>% 
  mutate(Klf4Diff = (KLF4 - RUNX1)) %>% 
  arrange(desc(Klf4Diff)) %>% 
  select(barcode, Sample, cellType, Klf4Diff) %>%
  slice_min(Klf4Diff, n = 5)
```


### Group by + sumarize : forget about loops

__First__: group by a particular variables
__Second__: summarize the data with new statistics.
__Summarize__: Turn many rows into one.

Examples:

- min(x) - minimum value of vector x.
- max(x) - maximum value of vector x.
- mean(x) - mean value of vector x.
- median(x) - median value of vector x.
- quantile(x, p) - pth quantile of vector x.
- sd(x) - standard deviation of vector x.
- var(x) - variance of vector x.
- IQR(x) - Inter Quartile Range (IQR) of vector x.
- diff(range(x)) - total range of vector x.

```{r}
d %>% 
  group_by(cellType) %>% 
  summarise(mean_total_counts = mean(total_counts, na.rm = TRUE), sd_total_counts = sd(total_counts), 
     mean_pct_Mt_count = mean(pct_counts_Mt), count = n()) %>% 
  #ungroup() %>% 
  slice_max(., n=20, order_by = mean_total_counts)  # note here, it does 
```

__Note: `mutate()` either changes an existing column or adds a new one. `summarise()` calculates a single value (per group). As you can see, in the first example, new column is added.__

```{r}
d %>% 
count(Sample, cellType)
```

# Plotting in R using `ggplot2`

`GGPlot2` is a powerful and a flexible R package, implemented by Hadley Wickham, for producing elegant graphics piece by piece (Wickham et al. 2017).

The `gg` in `ggplot2` means Grammar of Graphics, a graphic concept which describes plots by using a “grammar”. According to the ggplot2 concept, a plot can be divided into different fundamental parts: 

> Plot = Data + Aesthetics + Geometry


1. __Data:__ a data frame
2. __Aesthetics:__ used to indicate the x and y variables. It can be also used to control the color, the size and the shape of points, etc…..
3. __Geometry:__ corresponds to the type of graphics (scatter plot, histogram, box plot, line plot, ….)
4. additional layers for customization — title, labels, axis, etc.


## First plotting
The main function in the `ggplot2` package is `ggplot()`, which can be used to initialize the plotting system with __data__ and __x/y__ variables.

`ggplot()` creates a coordinate system to which you can add layers to. The first argument of `ggplot()` is the dataset to use in the graph. So `ggplot(data = d_exp)` creates an empty graph.

For example, the following R code takes the `KLF4` and `RUNX1` data set to initialize the `ggplot` and then a layer (geom_point()) is added onto the ggplot to create a scatter plot of x = KLF4 by y = RUNX1:

1. __Data=__ `d_exp`
2. __Aesthetic=:__ aes(x=KLF4, y=RUNX1)
3. __Geometry=__ `geom_point()`

```{r}
ggplot(d_exp)
```


```{r}
ggplot(d_exp, aes(x=KLF4, y=RUNX1))
```

```{r}
ggplot(d_exp, aes(x=KLF4, y=RUNX1)) + geom_point()
```

```{r}
ggplot(d_exp, aes(x=KLF4, y=RUNX1)) + geom_point(size = 1.2, color = "steelblue", shape = 21)
```

It’s also possible to control points shape and color by a grouping variable (here, `Sample`). For example, in the code below, we map points `color` and `shape` to the datasets grouping variable.

Note that, a `ggplot` can be holded in a variable, say `p`, to be printed later

```{r}
# Control points color by groups
ggplot(d_exp, aes(x=KLF4, y=RUNX1))+
  geom_point(aes(color = Sample))

# Change the default color manually.
# Use the scale_color_manual() function
p <- ggplot(d_exp, aes(x=KLF4, y=RUNX1))+
  geom_point(aes(color = Sample))+
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07", "#A2AFBB", "#17B8AB", "#3F4E77", "#FCFA27", "#BFAFFB", "#69B89B", "#7F4E97"))
print(p)
```

## GGPlot theme

Note that, the default theme of ggplots is `theme_gray()` (or `theme_grey()`), which is theme with grey background and white grid lines. More themes are available for professional presentations or publications. These include: `theme_bw()`, `theme_classic()` and `theme_minimal()`.

To change the theme of a given ggplot (p), use this: `p + theme_classic()`. 

```{r}
p <- ggplot(d_exp, aes(x=KLF4, y=RUNX1))+
  geom_point(aes(color = Sample))+
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07", "#A2AFBB", "#17B8AB", "#3F4E77", "#FCFA27", "#BFAFFB", "#69B89B", "#7F4E97"))
p <- p + theme_classic()
print(p)

```
```{r}
df <- reducedDim(sce)
head(df)
```

## Adding layers to ggplot, Lines (Prediction Line)
A plot constructed with ggplot can have more than one geom. In that case the mappings established in the `ggplot()` call are plot defaults that can be added to or overridden. Our plot could use a regression line:
```{r}
d_exp$pred.SC <- predict(lm(RUNX1 ~ KLF4, data = d_exp))

ggplot(d_exp, aes(x = KLF4, y = RUNX1)) + 
  geom_point(aes(color = Sample)) +
  geom_line(aes(y = pred.SC)) +
  theme_classic()
```


## `Title`, `xlab` & `ylab`
```{r}
df <- as.data.frame(reducedDim(sce))
df$Sample <- colData(sce)$Sample
p <- ggplot(df, aes(x=V1, y=V2))+
  geom_point(size = 0.4, aes(color = Sample))+
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07", "#A2AFBB", "#17B8AB", "#3F4E77", "#FCFA27", "#BFAFFB", "#69B89B", "#7F4E97")) + 
  ggtitle('t-SNE plot for Samples') + 
  xlab('tSNE-1') + 
  ylab('tSNE-2') + 
  theme_classic()

p
```

```{r}
df <- as.data.frame(reducedDim(sce))
df$Sample <- colData(sce)$Sample
p <- ggplot(df, aes(x=V1, y=V2))+
  geom_point(size = 0.4, aes(color = Sample))+
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07", "#A2AFBB", "#17B8AB", "#3F4E77", "#FCFA27", "#BFAFFB", "#69B89B", "#7F4E97")) + 
  ggtitle('t-SNE plot for Samples') + 
  xlab('tSNE-1') + 
  ylab('tSNE-2') + 
  theme_classic() + 
  guides(colour = guide_legend(override.aes = list(size=4)))

p
```
## Histogram

```{r}
ggplot(d_exp, aes(x=total_counts)) + geom_histogram() + theme_classic() 
```


## Density plot
```{r}
df <- data.frame(x=log10(sce$total_counts+1), Sample = sce$Sample)
ggplot(df,
       aes(x = x, fill = as.factor(Sample))) + 
       geom_density(alpha = 0.5) +
       labs(x = expression('log'[10]*'(Library Size)'), title = "Total reads density", fill = "Sample") + 
       theme_classic(base_size = 14) + # Setting the base size text for plots
       scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07", "#A2AFBB", "#17B8AB", "#3F4E77", "#FCFA27", "#BFAFFB", "#69B89B", "#7F4E97")) # Need to set fill manual
```

## Facet

```{r}
df <- data.frame(x=log10(sce$total_counts+1), Sample = sce$Sample)
ggplot(df,
       aes(x = x, fill = as.factor(Sample))) + 
       geom_density(alpha = 0.5) +
       labs(x = expression('log'[10]*'(Library Size)'), title = "Total reads density", fill = "Sample") + 
       theme_classic(base_size = 14) + # Setting the base size text for plots
       scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07", "#A2AFBB", "#17B8AB", "#3F4E77", "#FCFA27", "#BFAFFB", "#69B89B", "#7F4E97")) +  # Need to set fill manual 
  facet_wrap(~Sample)
```

# Statistical Transformations
## Statistical Transformations
Some plot types (such as scatterplots) do not require transformations–each point is plotted at x and y coordinates equal to the original value. Other plots, such as boxplots, histograms, prediction lines etc. require statistical transformations:

- for a smoother the y values must be transformed into predicted values

```{r}
ggplot(d_exp, aes(x = KLF4, y = RUNX1)) + 
  geom_point(aes(color = Sample)) +
  geom_smooth() +
  theme_classic()
```

```{r}
ggplot(d_exp, aes(x = KLF4, y = RUNX1)) + 
  geom_point(aes(color = Sample)) +
  geom_smooth(method = "lm") +
  theme_classic()
```

# Excercises
First let us look at the mpg dataset
```{r}
mpg
```


## Excercise #1
1. How would you plot the relationship between `cty`, the average city mileage, and `hwy`, the average highway mileage? How would you describe this relationship?


<details><summary>Click here for answer</summary>

```{r code1}
ggplot(mpg, aes(cty, hwy)) + 
  geom_point() +
  theme_classic()
```

</details>

<br>

2. How would you draw a smooth line? Also, how would you colour the transmissions differently?

<details><summary>Click here for answer</summary>
```{r code1_1}
ggplot(mpg, aes(cty, hwy)) + 
  geom_point(aes(colour = drv)) +
  geom_smooth(method = "lm") +
  theme_classic()
```

</details>


## Excercise #2
Show the distribution of scores for `hwy`, highway miles per gallon


<details><summary>Click here for answer</summary>

```{r code2}
ggplot(mpg, aes(hwy)) + 
  geom_histogram()
```

</details>

## Excercise #3
Produce a similar plot as shown below

```{r, echo=FALSE}
ggplot(mpg, aes(x=displ, y=cty, colour=drv, size=cyl)) +
  geom_point() + theme_classic()
```

<details><summary>Click here for answer</summary>

```{r code3}
ggplot(mpg, aes(x=displ, y=cty, colour=drv, size=cyl)) +
  geom_point() + theme_classic()
```

</details>


